META II \\ \large A Syntax-Oriented Compiler Writing Language

### 2017-01-10 doc/paper.tex
Two more typos in VALGOL I grammar transcription.
Transcribe VALGOL II compiler and example program.

IOST = 'EDIT' '(' EXP '.' .STRING
     .OUT('EDT' *) ')' /
     '.PRINT' .OUT('PNT') .,

@@

% TODO: continue transcribing Figures, starting from Figure 7 (VALGOL
% II Compiler written in META II).

>>>>

IOST = 'EDIT' '(' EXP ',' .STRING
     .OUT('EDT' *) ')' /
     'PRINT' .OUT('PNT') .,

@@

\begin{figure}
\caption{VALGOL II compiler written in META II}
\begin{verbatim}
.SYNTAX PROGRAM

ARRAYPART = '(.' EXP '.)' .OUT('AIA') .,

CALLPART = '(' .OUT('LDF') (EXP $(',' EXP) /
     .EMPTY) ')' .OUT('CLL') .,

VARIABLE = .ID .OUT('LD ' *) (ARRAYPART / .EMPTY) .,

PRIMARY = 'WHOLE' '(' EXP ')' .OUT('WHL') /
     .ID .OUT('LD ' *) (ARRAYPART / CALLPART / .EMPTY) /
     '.TRUE' .OUT('SET') / '.FALSE' .OUT('RST') /
     'O ' .OUT('RST') / '1 ' .OUT('SET') /
     .NUMBER .OUT('LDL' *) /
     '(' EXP ')' .,

TERM = PRIMARY $ ('*' PRIMARY .OUT('MLT') /
     '/' PRIMARY .OUT('DIV') /
     './.' PRIMARY .OUT('DIV') .OUT('WHL') ) .,

EXP2 = '-' TERM .OUT('NEG') /
     '+' TERM / TERM .,

EXP1 = EXP2 $('+' TERM .OUT('ADD') /
     '-' TERM .OUT('SUB')) .,

RELATION = EXP1 (
     '.L=' EXP1 .OUT('LEQ') /
     '.L' EXP1 .OUT('LES') /
     '.=' EXP1 .OUT('EQU') /
     '.-=' EXP1 .OUT('EQU') .OUT('NOT') /
     '.G=' EXP1 .OUT('LES') .OUT('NOT') /
     '.G' EXP1 .OUT('LEQ') .OUT('NOT') /
     .EMPTY) .,

BPRIMARY = '.-' RELATION .OUT('NOT') /
     RELATION .,

BTERM = BPRIMARY $ ('.,' .OUT('BF ' *1)
     .OUT('POP') BPRIAMRY)
     .LABEL *1 .,

BEXP1 = BTERM $( '.V' .OUT('BT ' *1)
     .OUT('POP') BTERM)
     .LABEL *1 .,

IMPLICATION1 = '.IMP' .OUT('NOT')
     .OUT('BT ' *1) .OUT('POP')
     BEXP1 .LABEL *1 .,

IMPLICATION = BEXP1 $ IMPLICATION1 .,

EQUIV = IMPLICATION $('.EQ' .OUT('EQU') ) .,

EXP = '.IF' EXP '.THEN' .OUT('BFP' *1)
     EXP .OUT('B  ' *2) .LABEL *1
     '.ELSE' EXP .LABEL *2 /
     EQUIV .,

ASSIGNPART = '=' EXP ( ASSIGNPART .OUT('ST') /
     ASSIGNPART / (CALLPART / .EMPTY
     .OUT('LDF') .OUT('CLL') )
     .OUT('POP') ) .,

WHILECLAUSE = '.WHILE' .OUT('BF ' *1)
     .OUT('POP') EXP .LABEL *1 / .EMPTY .,

FORCLAUSE = VARIABLE '=' .OUT('FLP')
     .OUT('BFP' *1) EXP '.STEP'
     .OUT('SST') .OUT('B  ' *2)
     .LABEL *1 EXP '.UNTIL' .OUT('ADS')
     .LABEL *2 .OUT('RSR') EXP
     .OUT('LEQ') WHILECLAUSE '.DO' .,

FORST = '.FOR' .OUT('SET') .LABEL *1
     FORCLAUSE .OUT('BFP' *2) ST
     .OUT('RST') .OUT('B  ' *1)
     .LABEL *2 .,

IOCALL = 'READ' '(' VARIABLE ',' EXP ')' .OUT('RED') /
     'WRITE' '(' VARIABLE ',' EXP ')' .OUT('WRT') /
     'EDIT' '(' EXP ',' .STRING
     .OUT('EDT' *) ')'
     'PRINT' .OUT('PNT') /
     'EJECT' .OUT('EJT') .,

IDSEQ1 = .ID .LABEL* .OUT('BLK 1') .,

TYPEDEC = '.REAL' IDSEQ .,

ARRAY1 = .ID .LABEL * '(.' 'O' '..' .NUMBER
     .OUT('BLK 1') .OUT('BLK' *) '.)' .,

ARRAYDEC = '.ARRAY' ARRAY1 $( ',' ARRAY1) .,

PROCEDURE = '.PROCEDURE' .ID .LABEL *
     .LABEL *1 .OUT('BLK 1') '('
     (IDSEQ / .EMPTY) ')' .OUT('SP  1') '.,'
     ST .OUT('R  ' *1) .,

DEC = TYPEDEC / ARRAYDEC / PROCEDURE .,

BLOCK = '.BEGIN' .OUT('B  ' *1) $(DEC '.,')
     .LABEL *1 ST $('.,' ST) '.END'
     (.ID / .EMPTY) .,

UNCONDITIONALST = IOCALL / ASSIGNCALLST /
     BLOCK .,

CONDST = '.IF' EXP '.THEN' .OUT('BFP' *1)
     (UNCONDITIONALST ('.ELSE' .OUT('B  ' *2)
     .LABEL *1 ST .LABEL *2 / .EMTPY
     .LABEL *1) / (FORST / UNTILST)
     .LABEL *1) .,

ST = CONDST / UNCONDITIONALST / FORST /
     UNTILST / .EMPTY .,

PROGRAM = BLOCK
     .OUT('HLT' .OUT('SP  1') .OUT('END') .,

% TODO: Figure (Order list of the VALGOL II machine)

\begin{figure}
\caption{Example program in VALGOL II}
\begin{verbatim}
.BEGIN
.PROCEDURE DETERMINANT(A, N) .,
.BEGIN
.PROCEDURE DUMP() .,
.BEGIN
.REAL D .,
.FOR D = 0 .STEP 1 .UNTIL N-1 .DO
     WRITE(MATRIX(. N*D .), N) .,
PRINT
.END DUMP .,
.PROCEDURE ABS(X) .,
     ABS = .IF 0 .L= X .THEN X .ELSE -X .,
.REAL PRODUCT, FACTOR, TEMP, R, I, J .,
PRODUCT = 1 .,
.FOR R = 0 .STEP 1 .UNTIL N-2
.WHILE PRODUCT .-= 0 .DO .BEGIN
     I  = R .,
     .FOR J = R+1 .STEP 1 .UNTIL N-1 .DO
          .IF ABS( A(. N*I + R .) ) .L
          ABS( A(. N*J + R .) ) .THEN
               I = J .,
     .IF A(. N*I + R .) .= 0 .THEN
          PRODUCT = 0
     .ELSE
          .IF I .-= R .THEN .BEGIN
               PRODUCT = -PRODUCT .,
               .FOR J = R .STEP 1 .UNTIL N-1 .DO
               .BEGIN
                    TEMP = A(. N*R + J .) .,
                    A(. N*R + J .) = A (. N* I + J .) .,
                    A(. N*I + J .) = TEMP .END .END .,
          TEMP = A(. N*R + R .) .,
          .FOR I = R+1 .STEP 1 .UNTIL N-1 .DO
          .BEGIN
               FACTOR = A(. N*I + R .) / TEMP .,
               .FOR J = R .STEP 1 .UNTIL N-1 .DO
                    A(. N*I + J .) = A(. N*I + J .)
                    -FACTOR * A(. N*R + J .) .,
          DUMP
          .END .END .,
.FOR I = 0 .STEP 1 .UNTIL N-1 .DO
     PRODUCT = PRODUCT * A(. N*I + I .) .,
DETERMINANT = PRODUCT
.END DETERMINANT .,
.REAL M, W, T ., .ARRAY MATRIX (. 0 .. 24 .) .,
.UNTIL .FALSE .DO .BEGIN
     EDIT(I, 'FIND DETERMINANT OF' ) ., PRINT., PRINT.,
     READ(M, 1) .,
     .FOR W = 0 .STEP 1 .UTNIL M-1 .DO .BEGIN
          READ(MATRIX (. M*W .), M) .,
          WRITE(MATRIX (. M*W .), M) .END .,
     PRINT .,  T = DETERMINANT (MATRIX, M) .,
     WRITE(T, 1) ., PRINT., PRINT .END
.END PROGRAM
\end{verbatim}
